---
title: 'Influence of inflammation on symptom heterogeneity in PTSD'
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
editor_options:
  markdown:
    wrap: 72
---

## Introduction

## Introduction

The aim of the simulation is to quantify the difference in symptom
heterogeneity (via Hill numbers) between two groups of individuals with
the same fictitious mental disorder: one group with and one without a
biological factor (e.g., inflammation). The disorder is defined by 5
symptom criteria of which at least two must be present to fulfill
diagnostic criteria. It is simulated with a binary latent construct "L",
present in 80% of the simulated individuals. The biological factor is
simulated with a second binary latent construct "I", present in 50% of
individuals and correlated with L (r = 0.6). The symptoms are simulated
as indicators "S1", "S2", "W1", "W2", and "M". "S" equates to a strong
indicator, "W" to a weak indicator, and "M" to a medium-strength
indicator. In this additive model, L and I each have a direct effect on
every symptom, with different strengths. Symptoms are measured on a
scale ranging from 0-4, with values of 2 or higher indicating symptom
presence. Symptom ratings are normally distributed around a mean of 2,
with more variance added for weaker indicators. Diagnosed individuals
are split into a group with (I = 1) and without (I = 0) the biological
factor. To ensure that differences in heterogeneity are not caused by
differences in severity, both groups are matched on the full
distribution of total symptom severity. Finally, the share of each
symptom combination and the heterogeneity (Hill numbers) are calculated
for both groups.

## Set-up

First, the necessary packages are loaded.

```{r library, echo=TRUE, message=FALSE, warning=FALSE}
# Load required packages
library(tidyverse)
library(MASS)
```

Second, a seed is set and the general set up is defined.

```{r general_setup, echo=TRUE}
# Setting seed for reproducibility
set.seed(123)

# All 32 possible combinations of the 5 binary symptom criteria
combinations <- expand.grid(c(0,1), c(0,1), c(0,1), c(0,1), c(0,1))

# Labels of the combinations, e.g. "11001" (order: S1, S2, W1, W2, M)
combination_labels <- apply(combinations, 1, paste0, collapse = "")
```

Third, a helper function is created that linearly rescales the simulated
values to a range from 0 to 4 (lowest value = 0, highest value = 4).

```{r helper_function_rescale, echo=TRUE}
# Function to ensure the values fall between 0 and 4
rescale <- function(x) {
  min_val <- min(x)
  max_val <- max(x)
  return((x - min_val) / (max_val - min_val) * 4)
}
```

Fourth, a function is created to generate correlated binary
variables

```{r helper_function_binary, echo=TRUE}
generate_correlated_binary <- function(n, p1, p2, cor) {
  # Draw n pairs of correlated standard normal values
  sigma <- matrix(c(1, cor, cor, 1), 2, 2)
  bvn <- mvrnorm(n, mu = c(0, 0), Sigma = sigma)
  
  # Dichotomize: values above the threshold become 1
  L <- ifelse(bvn[,1] > qnorm(1 - p1), 1, 0)
  I <- ifelse(bvn[,2] > qnorm(1 - p2), 1, 0)
  
  return(data.frame(L = L, I = I))
}
```

Fifth, a function is created to calculate Hill numbers. Hill numbers
quantify heterogeneity by measuring the effective number of categories
in a distribution. The parameter q controls sensitivity to rare
categories: q=0 counts all categories equally (richness), q=1 weights
by frequency (exponential of Shannon), and q=2 emphasizes common
categories (inverse Simpson).

```{r hill_function, echo=TRUE}
# Function to calculate Hill numbers
# Input: vector of probabilities, q parameter (order of diversity)
# Output: Hill number (effective number of categories)
calculate_hill <- function(probabilities, q) {
  # Only use probabilities > 0 (since log(0) is undefined for q=1)
  p <- probabilities[probabilities > 0]
  # Normalize to ensure probabilities sum to 1
  p <- p / sum(p)
  
  if (q == 1) {
    # Limit case: exponential of Shannon entropy
    return(exp(-sum(p * log(p))))
  } else {
    # General formula: Hill = (sum(p^q))^(1/(1-q))
    return(sum(p^q)^(1/(1-q)))
  }
}
```

Sixth, a function is created that calculates the share of each of the
32 symptom combinations in a group (number of individuals with the
combination divided by group size).

```{r combination_function, echo=TRUE}
calculate_combination_shares <- function(data) {
  # Code each individual's symptom combination, e.g. "11001"
  combination <- paste0(as.integer(data$S1_positive),
                        as.integer(data$S2_positive),
                        as.integer(data$W1_positive),
                        as.integer(data$W2_positive),
                        as.integer(data$M_positive))
  # Count each of the 32 combinations (including those with 0 individuals)
  counts <- table(factor(combination, levels = combination_labels))
  return(as.numeric(counts) / sum(counts))
}
```

Seventh, the set up of the simulation is defined.

```{r simulation_setup, echo=TRUE}
# Number of individuals and simulations
n <- 10000
n_sims <- 100

# Model parameters
p_L <- 0.8      # probability of the disorder (L)
p_I <- 0.5      # probability of the biological factor (I)
cor_LI <- 0.6   # correlation between L and I (underlying continuous values)

# Matrices to store the share of each symptom combination per simulation
results_WITH_inflammation <- matrix(0, ncol = 2^5, nrow = n_sims)
results_WITHOUT_inflammation <- matrix(0, ncol = 2^5, nrow = n_sims)

# Vectors to track sample sizes
n_WITH_inflammation <- numeric(n_sims)
n_WITHOUT_inflammation <- numeric(n_sims)

# Vectors to track symptom severity means
severity_WITH_inflammation <- numeric(n_sims)
severity_WITHOUT_inflammation <- numeric(n_sims)

# Vectors to track Hill numbers for heterogeneity analysis
# q=0: richness, q=1: exp(Shannon), q=2: inverse Simpson
hill_q0_WITH_inflammation <- numeric(n_sims)
hill_q0_WITHOUT_inflammation <- numeric(n_sims)
hill_q1_WITH_inflammation <- numeric(n_sims)
hill_q1_WITHOUT_inflammation <- numeric(n_sims)
hill_q2_WITH_inflammation <- numeric(n_sims)
hill_q2_WITHOUT_inflammation <- numeric(n_sims)
```

## Running Simulations

First, we simulate the presence of the mental disorder and inflammation status. Each individual has an 80% probability of having the disorder and a 50% probability of having inflammation, with a correlation of 0.6 between the two. Next, we generate individual symptom ratings for every simulated individual. These ratings are based on the previously stated assumptions, namely, they are distributed around a mean of 2 with varying standard deviations depending on symptom strength. Each symptom is influenced by both the disorder and inflammation through an additive model. After obtaining the symptom ratings, we introduce new variables that indicate whether a symptom is "present" or "absent". A symptom is deemed "present" if its rating is 2 or higher. We then evaluate the presence of the disorder based on the number of "present" symptoms. If at least 2 out of the 5 symptoms are deemed present, the individual is considered diagnosed with the disorder. The diagnosed individuals are then split into two groups: those with inflammation (I=1) and those without inflammation (I=0). To ensure that differences in heterogeneity are not caused by differences in severity, both groups are matched on the full
distribution of total symptom severity. Finally, the probability of each symptom combination is calculated separately for each inflammation group across 100 simulations.

```{r simulations, echo=TRUE}
for(sim in 1:n_sims) {
  # Generate correlated binary latent variables L (PTSD) and I (Inflammation)
  latent_vars <- generate_correlated_binary(n, p1 = p_L, p2 = p_I, cor = cor_LI)
  L <- latent_vars$L
  I <- latent_vars$I
  
# Simulate symptoms with BOTH L and I effects (additive model)
  # I factors are adjustable based on theoretical assumptions
  
  # Strong indicators - react strongly to PTSD, moderately to Inflammation
  S1 <- rescale(2.0 * L + 0.6 * I + rnorm(n, mean=2, sd=0.5))
  S2 <- rescale(1.8 * L + 0.5 * I + rnorm(n, mean=2, sd=0.5))
  
  # Medium indicator - reacts moderately to both
  M <- rescale(1.2 * L + 0.7 * I + rnorm(n, mean=2, sd=0.8))
  
  # Weak indicators - react weakly to PTSD, varied response to Inflammation
  W1 <- rescale(0.6 * L + 0.5 * I + rnorm(n, mean=2, sd=1))
  W2 <- rescale(0.4 * L + 0.3 * I + rnorm(n, mean=2, sd=1))
  
  
  # Combine into a data frame
  data <- data.frame(L, I, S1, S2, W1, W2, M)
  
  # Creating binary variables for all indicators
  data <- data %>%
    mutate(
      S1_positive = S1 >= 2,
      S2_positive = S2 >= 2,
      W1_positive = W1 >= 2,
      W2_positive = W2 >= 2,
      M_positive = M >= 2
    )
  
  # Calculate a new column that counts how many criteria are met for each individual
  data <- data %>%
    mutate(criteria_met = S1_positive + S2_positive + 
                          W1_positive + W2_positive + 
                          M_positive)
  
# Filter the dataset to keep only those individuals that meet at least 2 criteria
filtered_data <- data %>%
    filter(criteria_met >= 2)
  
# Calculate total symptom severity (sum of the five symptom ratings, range 0-20)
filtered_data <- filtered_data %>%
  mutate(severity = S1 + S2 + M + W1 + W2)

# Split into groups with (I = 1) and without (I = 0) the biological factor
filtered_WITH_I <- filtered_data %>% filter(I == 1)
filtered_WITHOUT_I <- filtered_data %>% filter(I == 0)

  # === Severity matching: equalize the full severity distribution ===
  # Severity is divided into bins of width 0.5; within each bin, the same
  # number of individuals is randomly drawn from both groups
  breaks <- seq(0, 20, by = 0.5)
  filtered_WITH_I <- filtered_WITH_I %>%
    mutate(bin = cut(severity, breaks, include.lowest = TRUE))
  filtered_WITHOUT_I <- filtered_WITHOUT_I %>%
    mutate(bin = cut(severity, breaks, include.lowest = TRUE))
  
  # Number of individuals to draw per bin (the smaller of the two groups)
  n_per_bin <- inner_join(count(filtered_WITH_I, bin, name = "n_with"),
                          count(filtered_WITHOUT_I, bin, name = "n_without"),
                          by = "bin") %>%
    mutate(n_take = pmin(n_with, n_without))
  
  # Randomly draw n_take individuals per bin from both groups
  filtered_WITH_I <- filtered_WITH_I %>%
    inner_join(n_per_bin, by = "bin") %>%
    group_by(bin) %>%
    filter(row_number() %in% sample(n(), first(n_take))) %>%
    ungroup()
  filtered_WITHOUT_I <- filtered_WITHOUT_I %>%
    inner_join(n_per_bin, by = "bin") %>%
    group_by(bin) %>%
    filter(row_number() %in% sample(n(), first(n_take))) %>%
    ungroup()

# Track sample sizes
n_WITH_inflammation[sim] <- nrow(filtered_WITH_I)
n_WITHOUT_inflammation[sim] <- nrow(filtered_WITHOUT_I)

# Track severity means (to verify matching worked)
severity_WITH_inflammation[sim] <- mean(filtered_WITH_I$severity)
severity_WITHOUT_inflammation[sim] <- mean(filtered_WITHOUT_I$severity)
  
  # Calculate the share of each symptom combination in both groups
  results_WITH_inflammation[sim, ] <- calculate_combination_shares(filtered_WITH_I)
  results_WITHOUT_inflammation[sim, ] <- calculate_combination_shares(filtered_WITHOUT_I)
  
  # Calculate Hill numbers (q = 0, 1, 2) for both groups
  hill_q0_WITH_inflammation[sim] <- calculate_hill(results_WITH_inflammation[sim, ], q = 0)
  hill_q1_WITH_inflammation[sim] <- calculate_hill(results_WITH_inflammation[sim, ], q = 1)
  hill_q2_WITH_inflammation[sim] <- calculate_hill(results_WITH_inflammation[sim, ], q = 2)
  hill_q0_WITHOUT_inflammation[sim] <- calculate_hill(results_WITHOUT_inflammation[sim, ], q = 0)
  hill_q1_WITHOUT_inflammation[sim] <- calculate_hill(results_WITHOUT_inflammation[sim, ], q = 1)
  hill_q2_WITHOUT_inflammation[sim] <- calculate_hill(results_WITHOUT_inflammation[sim, ], q = 2)
}
```

The simulation results are processed separately for each inflammation group. For both the "with inflammation" and "without inflammation" groups, we calculate the mean probability and standard deviation of each symptom combination across the 100 simulations. The two groups are then combined into a single dataset for comparison. Only symptom combinations that meet diagnostic criteria (≥2 symptoms present) are retained for analysis. Combinations are ranked by their average probability across both groups to facilitate visual comparison. Finally, we report the average sample sizes for each inflammation group across all simulations.

```{r calculations, echo=TRUE}
# Mean and SD of each combination's share across simulations, per group
derived_probs_WITH <- data.frame(
  Combination = combination_labels,
  Mean_Probability = apply(results_WITH_inflammation, 2, mean),
  SD_Probability = apply(results_WITH_inflammation, 2, sd),
  MeetsCriteria = rowSums(combinations) >= 2,
  Group = "With Inflammation")

derived_probs_WITHOUT <- data.frame(
  Combination = combination_labels,
  Mean_Probability = apply(results_WITHOUT_inflammation, 2, mean),
  SD_Probability = apply(results_WITHOUT_inflammation, 2, sd),
  MeetsCriteria = rowSums(combinations) >= 2,
  Group = "Without Inflammation")

# Combine both groups and keep only combinations that meet diagnostic criteria
filtered_probs_combined <- rbind(derived_probs_WITH, derived_probs_WITHOUT) %>%
  filter(MeetsCriteria) %>%
  arrange(Group, desc(Mean_Probability))

# Report average sample sizes
cat("Average sample sizes across simulations:\n")
cat("With Inflammation:", round(mean(n_WITH_inflammation), 1), "\n")
cat("Without Inflammation:", round(mean(n_WITHOUT_inflammation), 1), "\n")

# Report average symptom severity (to verify groups are matched)
cat("\nAverage symptom severity across simulations:\n")
cat("With Inflammation:", round(mean(severity_WITH_inflammation), 2), "\n")
cat("Without Inflammation:", round(mean(severity_WITHOUT_inflammation), 2), "\n")

# === Report Hill Numbers Results ===
cat("\n=== HETEROGENEITY ANALYSIS (Hill Numbers) ===\n")

# Calculate differences for each simulation
diff_q0 <- hill_q0_WITH_inflammation - hill_q0_WITHOUT_inflammation
diff_q1 <- hill_q1_WITH_inflammation - hill_q1_WITHOUT_inflammation
diff_q2 <- hill_q2_WITH_inflammation - hill_q2_WITHOUT_inflammation

# Create summary data frame
hill_summary <- data.frame(
  Measure = c("Hill q=0 (Richness)", 
              "Hill q=1 (exp Shannon)", 
              "Hill q=2 (inv Simpson)"),
  With_Inflammation_Mean = c(mean(hill_q0_WITH_inflammation),
                              mean(hill_q1_WITH_inflammation),
                              mean(hill_q2_WITH_inflammation)),
  With_Inflammation_SD = c(sd(hill_q0_WITH_inflammation),
                            sd(hill_q1_WITH_inflammation),
                            sd(hill_q2_WITH_inflammation)),
  Without_Inflammation_Mean = c(mean(hill_q0_WITHOUT_inflammation),
                                 mean(hill_q1_WITHOUT_inflammation),
                                 mean(hill_q2_WITHOUT_inflammation)),
  Without_Inflammation_SD = c(sd(hill_q0_WITHOUT_inflammation),
                               sd(hill_q1_WITHOUT_inflammation),
                               sd(hill_q2_WITHOUT_inflammation)),
  Difference = c(mean(diff_q0),
                 mean(diff_q1),
                 mean(diff_q2)),
  CI_95_Lower = c(mean(diff_q0) - 1.96 * sd(diff_q0) / sqrt(n_sims),
                  mean(diff_q1) - 1.96 * sd(diff_q1) / sqrt(n_sims),
                  mean(diff_q2) - 1.96 * sd(diff_q2) / sqrt(n_sims)),
  CI_95_Upper = c(mean(diff_q0) + 1.96 * sd(diff_q0) / sqrt(n_sims),
                  mean(diff_q1) + 1.96 * sd(diff_q1) / sqrt(n_sims),
                  mean(diff_q2) + 1.96 * sd(diff_q2) / sqrt(n_sims))
)

# Round all numeric columns
hill_summary[, 2:8] <- round(hill_summary[, 2:8], 3)

# Print summary table
print(hill_summary)
```

## Results

The table below shows the mean probability of each symptom combination that meets diagnostic
criteria, for both groups side by side. Combinations are sorted by their overall frequency (both groups combined), from most to least common; the last column shows the difference in mean probability.

```{r show_table, echo=TRUE}
# Mean share of each symptom combination per group (combinations meeting
# diagnostic criteria only), sorted by overall frequency
comparison_table <- data.frame(
  Combination = combination_labels,
  With_Inflammation = derived_probs_WITH$Mean_Probability,
  Without_Inflammation = derived_probs_WITHOUT$Mean_Probability) %>%
  filter(rowSums(combinations) >= 2) %>%
  mutate(Difference = With_Inflammation - Without_Inflammation) %>%
  arrange(desc(With_Inflammation + Without_Inflammation)) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

print(comparison_table)
```

## Visualization

The probabilities of symptom combinations are plotted for both inflammation groups. Each bar represents a symptom combination that meets diagnostic criteria (≥2 symptoms present). Orange bars represent individuals with inflammation, while blue bars represent individuals without inflammation. Combinations are ordered by their average probability across both groups. Error bars indicate the standard deviation across the 100 simulations.

```{r plot, echo=TRUE, message=FALSE, warning=FALSE}
# Define the theme for the figure
Theme_Figure_1a <- 
  theme(
    plot.title = element_text(size=12),
    axis.title.y = element_text(size=9, 
                                margin = margin(t = 0, r = 0, b = 0, l = 0)),
    axis.text.x = element_text(size = 7, angle = 90, hjust = 1, vjust = 0.5),
    axis.text.y = element_text(size=9, 
                               color = "black", 
                               margin = margin(t = 0, r = 0, b = 0, l = 5)),
    axis.ticks = element_blank(),
    panel.grid.major.x = element_blank(), 
    panel.grid.major.y = element_line(size=.2, 
                                      color="black"), 
    panel.grid.minor.y = element_blank(), 
    panel.border = element_rect(colour = "black", 
                                fill=NA, 
                                size=1),
    panel.background = element_rect(fill = "white"),
    legend.position = "bottom")

# Plot with two colors
ggplot(filtered_probs_combined, aes(x = reorder(Combination, -Mean_Probability), 
                                     y = Mean_Probability,
                                     fill = Group)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(aes(ymin = Mean_Probability - SD_Probability, 
                    ymax = Mean_Probability + SD_Probability), 
                position = position_dodge(width = 0.9),
                width = 0.25) +
  scale_fill_manual(values = c("With Inflammation" = "#F97134", 
                                "Without Inflammation" = "#3498db")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Symptom Combination Probabilities by Inflammation Status (Additive Model)", 
       x = "Symptom Combinations (S1, S2, W1, W2, M)", 
       y = "Probability",
       fill = "") + 
  Theme_Figure_1a
ggsave("Figure_2a_Additive_Model.png", width = 10, height = 6, dpi = 300)
```
