---
title: 'Influence of inflammation on symptom heterogeneity in PTSD'
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
editor_options:
  markdown:
    wrap: 72
---

## Introduction

The aim of the simulation is to calculate and compare the probability
distributions of symptom combinations between individuals with and
without inflammation in a fictitious mental disorder. The disorder is
defined by 5 symptom criteria of which at least 2 must be present to
fulfill diagnostic criteria. The disorder is simulated with a binary
latent variable L, which is present in 80% of the simulated individuals.
Five symptom indicators are simulated: S1, S2, M, W1, and W2, where S
stands for strong indicators, W for weak indicators, and M for a
medium-strength indicator. Each symptom is influenced by both the
disorder L and by a second binary latent variable I, which represents
inflammation. Inflammation is present in 50% of individuals. The
correlation between L and I is 0.6. Both latent variables act additively
on the symptoms, with varying strengths of effect. Symptoms are measured
on a scale from 0 to 4, with values of 2 or higher indicating symptom
presence. Symptom ratings are normally distributed around a mean of 2,
with different standard deviations depending on indicator strength. The
disorder is considered present if at least 2 of the 5 symptoms are
present. The simulation calculates the probability of each symptom
combination separately for individuals with and without inflammation.
To ensure fair comparison between inflammation groups, we simulate a large population and then select individuals from the non-inflammation group whose symptom severity matches the inflammation group. Specifically, we select non-inflammation individuals whose total symptom severity falls within ±0.5 SD of the inflammation group's mean severity, then randomly sample to achieve equal group sizes.

```{r setup, include=FALSE}
```

## Set-up

First, the necessary packages are loaded.

```{r library, echo=TRUE, message=FALSE, warning=FALSE}
# Load required packages
library(tidyverse)
library(mvtnorm)
library(MASS)
library(flextable)
library(officer)
```

Second, a seed is set and the general set up is defined.

```{r geenral_setup, echo=TRUE}
# Setting seed for reproducibility
set.seed(123)

# Combinations of binary criteria
combinations <- expand.grid(c(0,1), c(0,1), c(0,1), c(0,1), c(0,1))
```

Third, a helper function is created to assure that simulates values fall
between 0 and 4 (otherwise they are resampled).

```{r helper function, echo=TRUE}
# Function to ensure the values fall between 0 and 4
rescale <- function(x) {
  min_val <- min(x)
  max_val <- max(x)
  return((x - min_val) / (max_val - min_val) * 4)
}
```

Fourth, a function is created to generate correlated binary
variables

```{r helper function, echo=TRUE}
generate_correlated_binary <- function(n, p1, p2, cor) {
  means <- c(0, 0)
  sigma <- matrix(c(1, cor, cor, 1), 2, 2)
  bvn <- mvrnorm(n, mu = means, Sigma = sigma)
  
  threshold_L <- qnorm(1 - p1)
  threshold_I <- qnorm(1 - p2)
  
  L <- ifelse(bvn[,1] > threshold_L, 1, 0)
  I <- ifelse(bvn[,2] > threshold_I, 1, 0)
  
  return(data.frame(L = L, I = I))
}
```

Fifth, the set up of the simulation is defined.

```{r simulation_setup, echo=TRUE}
# Number of individuals and simulations
n <- 10000
n_sims <- 100

# Matrices to accumulate the results of each simulation
results_WITH_inflammation <- matrix(0, ncol = 2^5, nrow = n_sims)
results_WITHOUT_inflammation <- matrix(0, ncol = 2^5, nrow = n_sims)

# Vectors to track sample sizes
n_WITH_inflammation <- numeric(n_sims)
n_WITHOUT_inflammation <- numeric(n_sims)

# Vectors to track symptom severity means
severity_WITH_inflammation <- numeric(n_sims)
severity_WITHOUT_inflammation <- numeric(n_sims)
```

## Running Simulations

First, we simulate the presence of the mental disorder and inflammation status. Each individual has an 80% probability of having the disorder and a 50% probability of having inflammation, with a correlation of 0.6 between the two. Next, we generate individual symptom ratings for every simulated individual. These ratings are based on the previously stated assumptions, namely, they are distributed around a mean of 2 with varying standard deviations depending on symptom strength. Each symptom is influenced by both the disorder and inflammation through an additive model. After obtaining the symptom ratings, we introduce new variables that indicate whether a symptom is "present" or "absent". A symptom is deemed "present" if its rating is 2 or higher. We then evaluate the presence of the disorder based on the number of "present" symptoms. If at least 2 out of the 5 symptoms are deemed present, then the disorder is considered to be "present" in that individual. The diagnosed individuals are then split into two groups: those with inflammation (I=1) and those without inflammation (I=0). Finally, the probability of each symptom combination is calculated separately for each inflammation group across 100 simulations.

```{r simulations, echo=TRUE}
for(sim in 1:n_sims) {
  # NEW: Generate correlated binary latent variables L (PTSD) and I (Inflammation)
  latent_vars <- generate_correlated_binary(n, p1 = 0.8, p2 = 0.5, cor = 0.6)
  L <- latent_vars$L
  I <- latent_vars$I
  
# NEW: Simulate symptoms with BOTH L and I effects (additive model)
  # I factors are adjustable based on theoretical assumptions
  
  # Strong indicators - react strongly to PTSD, moderately to Inflammation
  S1 <- rescale(2.0 * L + 0.6 * I + rnorm(n, mean=2, sd=0.5))
  S2 <- rescale(1.8 * L + 0.5 * I + rnorm(n, mean=2, sd=0.5))
  
  # Medium indicator - reacts moderately to both
  M <- rescale(1.2 * L + 0.7 * I + rnorm(n, mean=2, sd=0.8))
  
  # Weak indicators - react weakly to PTSD, varied response to Inflammation
  W1 <- rescale(0.6 * L + 0.5 * I + rnorm(n, mean=2, sd=1))
  W2 <- rescale(0.4 * L + 0.3 * I + rnorm(n, mean=2, sd=1))
  
  
  # Combine into a data frame
  data <- data.frame(L, I, S1, S2, W1, W2, M)
  
  # Creating binary variables for all indicators
  data <- data %>%
    mutate(
      S1_positive = S1 >= 2,
      S2_positive = S2 >= 2,
      W1_positive = W1 >= 2,
      W2_positive = W2 >= 2,
      M_positive = M >= 2
    )
  
  # Calculate a new column that counts how many criteria are met for each individual
  data <- data %>%
    mutate(criteria_met = S1_positive + S2_positive + 
                          W1_positive + W2_positive + 
                          M_positive)
  
# Filter the dataset to keep only those individuals that meet at least 2 criteria
filtered_data <- data %>%
    filter(criteria_met >= 2)
  
# Calculate symptom severity (sum of all 5 symptom ratings)
filtered_data <- filtered_data %>%
  mutate(severity = S1 + S2 + M + W1 + W2)

# Split into inflammation groups
filtered_WITH_I <- filtered_data %>% filter(I == 1)
filtered_WITHOUT_I <- filtered_data %>% filter(I == 0)

# === Severity-based sampling for equal groups ===
# Calculate mean and SD of severity in inflammation group
mean_severity_WITH <- mean(filtered_WITH_I$severity)
sd_severity_WITH <- sd(filtered_WITH_I$severity)

# Define selection range: mean ± 0.5 SD
lower_bound <- mean_severity_WITH - 0.5 * sd_severity_WITH
upper_bound <- mean_severity_WITH + 0.5 * sd_severity_WITH

# Select non-inflammation individuals within this severity range
filtered_WITHOUT_I_eligible <- filtered_WITHOUT_I %>%
  filter(severity >= lower_bound & severity <= upper_bound)

# Determine sample size (minimum of both groups)
n_sample <- min(nrow(filtered_WITH_I), nrow(filtered_WITHOUT_I_eligible))

# Random sample from both groups to get equal sizes
if(n_sample > 5) {
  filtered_WITH_I <- filtered_WITH_I %>% sample_n(n_sample)
  filtered_WITHOUT_I <- filtered_WITHOUT_I_eligible %>% sample_n(n_sample)
}

# Track sample sizes
n_WITH_inflammation[sim] <- nrow(filtered_WITH_I)
n_WITHOUT_inflammation[sim] <- nrow(filtered_WITHOUT_I)

# Track severity means (to verify matching worked)
severity_WITH_inflammation[sim] <- mean(filtered_WITH_I$severity)
severity_WITHOUT_inflammation[sim] <- mean(filtered_WITHOUT_I$severity)
  
# === Calculate for WITH inflammation group ===
if(nrow(filtered_WITH_I) > 5) {  # Need sufficient data for covariance
    cov_matrix_WITH <- cov(filtered_WITH_I[, c("S1", "S2", "W1", "W2", "M")])
    means_WITH <- colMeans(filtered_WITH_I[, c("S1", "S2", "W1", "W2", "M")])
    
    for(i in 1:nrow(combinations)){
      lower <- as.numeric(ifelse(combinations[i,] == 1, 2, -Inf))
      upper <- as.numeric(ifelse(combinations[i,] == 1, Inf, 2))
      results_WITH_inflammation[sim, i] <- pmvnorm(lower = lower, 
                                                   upper = upper, 
                                                   mean = means_WITH, 
                                                   sigma = cov_matrix_WITH)
    }
}
  
# === Calculate for WITHOUT inflammation group ===
if(nrow(filtered_WITHOUT_I) > 5) {  # Need sufficient data for covariance
    cov_matrix_WITHOUT <- cov(filtered_WITHOUT_I[, c("S1", "S2", "W1", "W2", "M")])
    means_WITHOUT <- colMeans(filtered_WITHOUT_I[, c("S1", "S2", "W1", "W2", "M")])
    
    for(i in 1:nrow(combinations)){
      lower <- as.numeric(ifelse(combinations[i,] == 1, 2, -Inf))
      upper <- as.numeric(ifelse(combinations[i,] == 1, Inf, 2))
      results_WITHOUT_inflammation[sim, i] <- pmvnorm(lower = lower, 
                                                      upper = upper, 
                                                      mean = means_WITHOUT, 
                                                      sigma = cov_matrix_WITHOUT)
    }
  }
}
```

The simulation results are processed separately for each inflammation group. For both the "with inflammation" and "without inflammation" groups, we calculate the mean probability and standard deviation of each symptom combination across the 100 simulations. The two groups are then combined into a single dataset for comparison. Only symptom combinations that meet diagnostic criteria (≥2 symptoms present) are retained for analysis. Combinations are ranked by their average probability across both groups to facilitate visual comparison. Finally, we report the average sample sizes for each inflammation group across all simulations.

```{r calculations, echo=TRUE}
# === Process results for WITH inflammation group ===
derived_probs_WITH <- data.frame(
  Combination = paste(combinations[,1], 
                      combinations[,2], 
                      combinations[,3], 
                      combinations[,4], 
                      combinations[,5], sep=""),
  Mean_Probability = apply(results_WITH_inflammation, 2, mean),
  SD_Probability = apply(results_WITH_inflammation, 2, sd),
  Group = "With Inflammation")

# === Process results for WITHOUT inflammation group ===
derived_probs_WITHOUT <- data.frame(
  Combination = paste(combinations[,1], 
                      combinations[,2], 
                      combinations[,3], 
                      combinations[,4], 
                      combinations[,5], sep=""),
  Mean_Probability = apply(results_WITHOUT_inflammation, 2, mean),
  SD_Probability = apply(results_WITHOUT_inflammation, 2, sd),
  Group = "Without Inflammation")

# Combine both groups
derived_probs_combined <- rbind(derived_probs_WITH, derived_probs_WITHOUT)

# Determine which combinations meet 2 out of 5 criteria
derived_probs_combined$MeetsCriteria <- ifelse(rowSums(combinations) >= 2, TRUE, FALSE)

# Filter to only include combinations that meet the criteria
filtered_probs_combined <- derived_probs_combined %>% 
  filter(MeetsCriteria) %>%
  arrange(Group, desc(Mean_Probability))

# Add abbreviations for x-axis
# We'll rank combinations by their average probability across both groups
avg_probs <- derived_probs_combined %>%
  filter(MeetsCriteria) %>%
  group_by(Combination) %>%
  summarise(Avg = mean(Mean_Probability)) %>%
  arrange(desc(Avg))

filtered_probs_combined$Rank <- match(filtered_probs_combined$Combination, avg_probs$Combination)

# Report average sample sizes
cat("Average sample sizes across simulations:\n")
cat("With Inflammation:", round(mean(n_WITH_inflammation), 1), "\n")
cat("Without Inflammation:", round(mean(n_WITHOUT_inflammation), 1), "\n")

# Report average symptom severity (to verify groups are matched)
cat("\nAverage symptom severity across simulations:\n")
cat("With Inflammation:", round(mean(severity_WITH_inflammation), 2), "\n")
cat("Without Inflammation:", round(mean(severity_WITHOUT_inflammation), 2), "\n")
```

## Results

The tables below showcase the mean probabilities and standard deviations for all symptom combinations that meet diagnostic criteria, displayed separately for individuals with and without inflammation. Combinations are sorted by probability within each group, from most to least common.

```{r show_table, echo=TRUE}
# Add MeetsCriteria column to individual dataframes
derived_probs_WITH$MeetsCriteria <- ifelse(rowSums(combinations) >= 2, TRUE, FALSE)
derived_probs_WITHOUT$MeetsCriteria <- ifelse(rowSums(combinations) >= 2, TRUE, FALSE)

# Show ALL results for WITH inflammation
cat("\n=== WITH INFLAMMATION (all combinations meeting criteria) ===\n")
print(derived_probs_WITH %>% 
        filter(MeetsCriteria) %>% 
        arrange(desc(Mean_Probability)))

# Show ALL results for WITHOUT inflammation
cat("\n=== WITHOUT INFLAMMATION (all combinations meeting criteria) ===\n")
print(derived_probs_WITHOUT %>% 
        filter(MeetsCriteria) %>% 
        arrange(desc(Mean_Probability)))
```

## Visualization

The probabilities of symptom combinations are plotted for both inflammation groups. Each bar represents a symptom combination that meets diagnostic criteria (≥2 symptoms present). Orange bars represent individuals with inflammation, while blue bars represent individuals without inflammation. Combinations are ordered by their average probability across both groups. Error bars indicate the standard deviation across the 100 simulations.

```{r plot, echo=TRUE, message=FALSE, warning=FALSE}
# Define the theme for the figure
Theme_Figure_1a <- 
  theme(
    plot.title = element_text(size=12),
    axis.title.y = element_text(size=9, 
                                margin = margin(t = 0, r = 0, b = 0, l = 0)),
    axis.text.x = element_text(size = 7, angle = 90, hjust = 1, vjust = 0.5),
    axis.text.y = element_text(size=9, 
                               color = "black", 
                               margin = margin(t = 0, r = 0, b = 0, l = 5)),
    axis.ticks = element_blank(),
    panel.grid.major.x = element_blank(), 
    panel.grid.major.y = element_line(size=.2, 
                                      color="black"), 
    panel.grid.minor.y = element_blank(), 
    panel.border = element_rect(colour = "black", 
                                fill=NA, 
                                size=1),
    panel.background = element_rect(fill = "white"),
    legend.position = "bottom")

# Plot with two colors
ggplot(filtered_probs_combined, aes(x = reorder(Combination, -Mean_Probability), 
                                     y = Mean_Probability,
                                     fill = Group)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(aes(ymin = Mean_Probability - SD_Probability, 
                    ymax = Mean_Probability + SD_Probability), 
                position = position_dodge(width = 0.9),
                width = 0.25) +
  scale_fill_manual(values = c("With Inflammation" = "#F97134", 
                                "Without Inflammation" = "#3498db")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Symptom Combination Probabilities by Inflammation Status (Additive Model)", 
       x = "Symptom Combinations (S1, S2, W1, W2, M)", 
       y = "Probability",
       fill = "") + 
  Theme_Figure_1a
```
