---
title: 'Influence of inflammation on symptom heterogeneity in PTSD'
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
editor_options:
  markdown:
    wrap: 72
---

## Introduction

The aim of the simulation is to calculate the probability of all symptom
combinations of a fictitious mental disorder. The disorder is considered
unifaceted and is defined by 5 symptom criteria of which at least two
must be present to fulfill diagnostic criteria. The disorder is
simulated with a binary (present or absent) latent construct "L". The
symptoms are simulated as indicators "S1", "S2", "W1", "W2", and "M".
"S" equates to a strong indicator, "W" to a weak indicator, and "M" to a
medium-strength indicator. The mental disorder is assumed to be present
in half of the simulated individuals. Symptoms are measured on a scale
ranging from 0-4, with values of 2 or higher indicating symptom
presence. Symptom ratings are assumed to be normally distributed around
a mean of 2, but with different standard deviations across the simulated
population. Depending on their "strength", additional variance is added
(with more variance added for weaker indicator strength). The presence
of the disorder is inferred if at least 2 of the 5 criteria (symptoms)
are met. Finally, the probabilities of each combination are inferred,
shown, and plotted. **Ich habe die Simulation mit einer zweiten latenten
Variable I, welche Inflammation simuliert, erweitert. Die Variable ist
binär und bei der Hälfte der Individuen vorhanden. Zwischen I und und
der latenten Variable L, welche die Krankheit simuliert, besteht eine
Korrelation von 0.6. Zwischen I und den Indikatoren S1, S2, M, W1 und W2
bestehen unterschiedlich starke Zusammenhänge. Das Modell ist so
gewählt, dass beide latenten Variablen additiv, das heisst unabhängig,
auf die Indikatoren wirken.**

```{r setup, include=FALSE}
```

## Set-up

First, the necessary packages are loaded.

```{r library, echo=TRUE, message=FALSE, warning=FALSE}
# Load required packages
install.packages("mvtnorm")
library(tidyverse)
library(mvtnorm)
library(MASS)
library(flextable)
library(officer)
```

Second, a seed is set and the general set up is defined.

```{r geenral_setup, echo=TRUE}
# Setting seed for reproducibility
set.seed(123)

# Combinations of binary criteria
combinations <- expand.grid(c(0,1), c(0,1), c(0,1), c(0,1), c(0,1))
```

Third, a helper function is created to assure that simulates values fall
between 0 and 4 (otherwise they are resampled).

```{r helper function, echo=TRUE}
# Function to ensure the values fall between 0 and 4
rescale <- function(x) {
  min_val <- min(x)
  max_val <- max(x)
  return((x - min_val) / (max_val - min_val) * 4)
}
```

NEW: Fourth, a function is created to generate correlated binary
variables

```{r helper function, echo=TRUE}
generate_correlated_binary <- function(n, p1, p2, cor) {
  means <- c(0, 0)
  sigma <- matrix(c(1, cor, cor, 1), 2, 2)
  bvn <- mvrnorm(n, mu = means, Sigma = sigma)
  
  threshold_L <- qnorm(1 - p1)
  threshold_I <- qnorm(1 - p2)
  
  L <- ifelse(bvn[,1] > threshold_L, 1, 0)
  I <- ifelse(bvn[,2] > threshold_I, 1, 0)
  
  return(data.frame(L = L, I = I))
}
```

Fifth, the set up of the simulation is defined.

```{r simulation_setup, echo=TRUE}
# Number of individuals and simulations
n <- 1000
n_sims <- 100

# Matrix to accumulate the results of each simulation
results_accumulator <- matrix(0, ncol = 2^5, nrow = n_sims)
```

## Running Simulations

First, we simulate the presence of the "mental disorder". Each
individual has a 50% probability of having the disorder. Next, we
generate individual symptom ratings for every simulated individual.
These ratings are based on the previously stated assumptions, namely,
they are distributed around a mean of 2 with varying standard deviations
depending on symptom strength. After obtaining the symptom ratings, we
introduce new variables that indicate whether a symptom is "present" or
"absent". A symptom is deemed "present" if its rating is 2 or higher. We
then evaluate the presence of the disorder based on the number of
"present" symptoms. If at least 2 out of the 5 symptoms are deemed
present, then the disorder is considered to be "present" in that
individual. Lastly, we calculate the probability of each combination of
symptoms.

```{r simulations, echo=TRUE}
for(sim in 1:n_sims) {
  # NEW: Generate correlated binary latent variables L (PTSD) and I (Inflammation)
  latent_vars <- generate_correlated_binary(n, p1 = 0.8, p2 = 0.5, cor = 0.6)
  L <- latent_vars$L
  I <- latent_vars$I
  
# NEW: Simulate symptoms with BOTH L and I effects (additive model)
  # I factors are adjustable based on theoretical assumptions
  
  # Strong indicators - react strongly to PTSD, moderately to Inflammation
  S1 <- rescale(2.0 * L + 1.2 * I + rnorm(n, mean=2, sd=0.5))
  S2 <- rescale(1.8 * L + 1.0 * I + rnorm(n, mean=2, sd=0.5))
  
  # Medium indicator - reacts moderately to both
  M <- rescale(1.2 * L + 1.3 * I + rnorm(n, mean=2, sd=0.8))
  
  # Weak indicators - react weakly to PTSD, varied response to Inflammation
  W1 <- rescale(0.6 * L + 0.9 * I + rnorm(n, mean=2, sd=1))
  W2 <- rescale(0.4 * L + 0.6 * I + rnorm(n, mean=2, sd=1))
  
  
  # Combine into a data frame
  data <- data.frame(L, S1, S2, W1, W2, M)
  
  # Creating binary variables for all indicators
  data <- data %>%
    mutate(
      S1_positive = S1 >= 2,
      S2_positive = S2 >= 2,
      W1_positive = W1 >= 2,
      W2_positive = W2 >= 2,
      M_positive = M >= 2
    )
  
  # Calculate a new column that counts how many criteria are met for each individual
  data <- data %>%
    mutate(criteria_met = S1_positive + S2_positive + 
                          W1_positive + W2_positive + 
                          M_positive)
  
  # Filter the dataset to keep only those individuals that meet at least 2 criteria
  filtered_data <- data %>%
    filter(criteria_met >= 2)
  
  # Estimate covariance matrix
  cov_matrix <- cov(filtered_data[, c("S1", "S2", "W1", "W2", "M")])
  means_simulated <- colMeans(filtered_data[, c("S1", "S2", "W1", "W2", "M")])
  
  for(i in 1:nrow(combinations)){
    lower <- as.numeric(ifelse(combinations[i,] == 1, 2, -Inf))
    upper <- as.numeric(ifelse(combinations[i,] == 1, Inf, 2))
    results_accumulator[sim, i] <- pmvnorm(lower = lower, 
                                           upper = upper, 
                                           mean = means_simulated, 
                                           sigma = cov_matrix)
  }
}
```

Then, the results are newly organized and labelled correctly.

```{r calculations, echo=TRUE}
# Compute the mean and SD for each combination
derived_probs <- data.frame(
  Combination = paste(combinations[,1], 
                      combinations[,2], 
                      combinations[,3], 
                      combinations[,4], 
                      combinations[,5], sep=""),
                            Mean_Probability = apply(results_accumulator, 2, mean),
                            SD_Probability = apply(results_accumulator, 2, sd))

# Determine which combinations meet 2 out of 5 criteria
derived_probs$MeetsCriteria <- ifelse(rowSums(combinations) >= 2, TRUE, FALSE)

# Sort by Mean_Probability
derived_probs <- derived_probs %>%
  arrange(desc(Mean_Probability))

# Add abbreviations to the results dataframe for cleaner x-axis labels
derived_probs$Abbreviation <- as.character(1:nrow(derived_probs))

# Filter derived_probs to only include combinations that meet the criteria
filtered_probs <- derived_probs %>% 
  filter(MeetsCriteria)
```

## Results

The table below showcases the mean probabilities and standard deviations
for various criteria combinations:

```{r show_table, echo=TRUE}
print(derived_probs)
```

## Visualization

The probabilities of the combinations that are indicative of disorder
presence are plotted.

```{r plot, echo=TRUE, message=FALSE, warning=FALSE}
# Define the theme for the figure
Theme_Figure_1a <- 
  theme(
    plot.title = element_text(size=12),
    axis.title.y = element_text(size=9, 
                                margin = margin(t = 0, r = 0, b = 0, l = 0)),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size=9, 
                               color = "black", 
                               margin = margin(t = 0, r = 0, b = 0, l = 5)),
    axis.ticks = element_blank(),
    panel.grid.major.x = element_blank(), 
    panel.grid.major.y = element_line(size=.2, 
                                      color="black" ), 
    panel.grid.minor.y = element_blank(), 
    panel.border = element_rect(colour = "black", 
                                fill=NA, 
                                size=1),
    panel.background = element_rect(fill = "white"))

# Plot 
ggplot(filtered_probs, aes(x = reorder(Abbreviation, -Mean_Probability), 
                                 y = Mean_Probability)) +
  geom_bar(stat = "identity", fill = "#F97134") +
  geom_errorbar(aes(ymin = Mean_Probability - SD_Probability, 
                    ymax = Mean_Probability + SD_Probability), 
                width = 0.25) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Simulation", x = NULL, y = "Probability") + 
  theme(legend.position = "none") +  
  Theme_Figure_1a
```
